### 2016-10-11 robust-software.tex
fixed some spacing and leftover elements
Smoothed out words, added build utility intro, added two more suggested references, spell-checked

\begin{itemize}

\item
References:
  \begin{itemize}
  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-2-15
  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-3-31
  \item https://tspace.library.utoronto.ca/handle/1807/73111
  \end{itemize}

\item
Honor local conventions,
e.g.,
use \texttt{TMPDIR} for temporary files.
(The problem is,
where are all of these written down?)

\item
Combine versioning and making previous versions available into a single point
in order to free up space for something else to be rule 10?

\end{itemize}

@@

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. With a lack of
robustness also comes a lack of reproducibility and many duplicated efforts,
slowing the pace of research~\cite{prabhu2011,lawlor2015}.
Bioinformatics is full of such efforts, catalogued by repositories such as
the ELIXIR tools and data registry~\cite{ison2016}
and the Bioinformatics Links Directory~\cite{brazas2012}. In the Bioinformatics
Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.
Some of these tools are legitimate efforts to improve the state-of-the-art, but
often they are difficult to install and run~\cite{stajich2002}, and are effectively abandoned
after publication~\cite{nekrutenko2012}.

@@

\textbf{List required dependencies:} Often, software depends on
specific versions of libraries, modules, or operating systems. 
This is
entirely reasonable as long as it is properly documented. Include the name and
version number for each dependency.  Several methods exist for acquiring
required packages: either provide the commands necessary to download the
dependencies; link to the software homepage; or (our preference) use a
dependency manager. We address dependencies in more detail in rule
\textcolor{red}{5}.

\textbf{Installation instructions:} If the software needs to be compiled
or installed, list those instructions in the README. Also
mention if you recommend they use a pre-compiled binary instead
through a system such as pip, yum or apt.

\textbf{Input and output files}: All possible input and output files
should be listed in this section. Do the files conform to a particular
standard, an extension of an existing format, or is it your own
format? If using a standard format, link to the specification and
version. If you extend the standard or have your own format, define it
explicitly, listing all the required fields and acceptable values.
(You get bonus points if you include a script to convert between
a standard format and your file format). If there is no rigorous format
(such as with log files), show an example file, or the first few lines,
and explain what the sections mean.

Input files and their formats are included in most documentation, but
intermediate, auxilliary and log files are often missing. 
\emph{All} files should be listed in the README, even those considered self-explanatory. Log
files are often full of valuable information that can be
mined for the user's specific purpose. If your users might need to know,
``Does this program report the percentage of reads trimmed to remove
adapter sequences?'' they should be able to check the README and confidently
say, ``Yes, it is in the log file''.

\textbf{Attributions and licensing:} Attributions are how you credit
your main contributors; licenses are how you want others to use and
credit your software. READMEs are usualy available \emph{before}
the software is installed (or even downloaded).
Leave no
question in anyone's mind about whether your software can be used
commercially, how much modification is permitted, and how other software
needs to credit you. If your software is not open source, state that clearly.
Attributions can also contain a list of ``expert'' users
that can be contacted if new users have problems with the software.

@@

Usage should provide all of the information necessary to run the
software. It is invoked either by running the software without
any arguments; running the software with incorrect arguments; or by
explicitly choosing a help or usage option.

@@

The version of your software should be easily available.
Supplying \texttt{-\/-version} or \texttt{-v} on the command line should print
the software name and version number, and it should
also be included in all of the program's output, particularly debugging
traces: if someone needs help, it's important that they be able to tell
whoever's helping them which version of the software they're using.

While new releases may make a program better in general,
they can simultaneously create work for someone
who integrated the old version into their own workflow a year or two ago,
and won't see any benefits from upgrading.
A program's authors should therefore ensure that every version they have released
continues to be available. A number of mechanisms exist for
controlled release that range from as simple as adding an appropriate
commit message or tag to version control, to official releases alongside
code on Sourceforge, Bitbucket or GitHub, to depositing into a
repository like apt, yum, homebrew, CPAN, etc. Choose the method that
best suits the number and expertise of users you anticipate.

@@

Unfortunately, reusing software introduces dependencies, which can
bring their own special pain. The interface between two software
packages can also be a source of considerable frustration: all too
often, support requests descend into debugging errors produced by the
other project. In addition, every package someone has to install
before being able to use yours is a
possible (some would say ``likely'') source of frustration for some
potential user. The way the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example, GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

@@

Second, \textbf{ensure the appropriate software and version is available}.
Early in execution, check whether the dependency is executable and whether the
version is compatible.
Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using a
dependency management system (see Rule \textcolor{red}{RULENO}).

@@

\section{Use a build system and dependency manager.}

If the software depends on a build tool, either because it is written in a
compiled language or to manage a workflow, it should use a standard tool such
as Make rather than home-brewed scripts.

To strike a balance between
these two, developers should
\textbf{document \emph{all} of the packages that theirs depends on, preferably in a machine-readable form}.
For example, it is common for Python projects to include a file called
\texttt{requirements.txt} that lists the names of required libraries,
along with version ranges:

@@

This file can be read by a package manager, which can check that the
required software is available, and install it if it is not. Similar
mechanisms exist for Perl, R, and other languages.  Whatever is used,
we recommend that developers should \emph{always} install dependencies
using this mechanism, especially on their personal machines, so that
they're sure it works.

Conversely, developers should
\textbf{avoid depending on scripts and tools which are not available as packages},
even (or especially) if they are installed on the computers the original developer is using.
In many cases, a program's author may not realize that some tool was built locally, and
doesn't exist elsewhere. At present, the only sure way to discover such
unknown dependencies is to install on a system administered by someone
else and see what breaks. As use of lightweight
virtualization containers like Docker becomes more widespread, it may
become common to test installation on a virtual machine.

@@

Testing the ability to install software has traditionally been regarded as difficult,
since it necessarily alters the machine on which the test is conducted.
Lightweight virtualization containers like Docker make this much easier as
well, or simply asking another person to try and build your software.

@@

\textbf{Allow the user to configure useful parameters on the command line.}
Providing parameters on the
command line increases the flexibility and usability of the program. You
may have determined early in development that 0.58 is an optimal seed for your
original dataset, but that doesn't mean that is the best seed for every
case. Being able to change parameters on the fly to determine if and how
they change the results is important as your software gains more users,
facilitating exploratory analysis and parameter sweeping.

When the software starts, it should \textbf{echo all parameters and software
versions to standard out or a log file alongside the results}. This
feature supports greater reproducibility because any result can be
replicated with only the previous output files as reference.

You can set reasonable default values
as long as any command line arguments
override those values. Configuration files should be used in preference
to hardcoding the defaults directly. Configuration files can be in a standard
location, e.g. \texttt{.packagerc} in the user's home directory or
provided on the command line as an additional argument.
Only values that are unlikely to
change between runs belong in the config file, such as dependencies,
servers, version numbers, network drives, and any other defaults for
your lab or institution. 
Since
all parameters should be echoed in the results, config files should be
cleaned up after execution completes or they occupy valuable disk space.
Use a default configuration file and specify all other parameters on the
command line.

Finally,
if the software is configurable in any way,
it should \textbf{check that all input values are in a reasonable range},
\textbf{choose reasonable defaults where they exist},
and \textbf{set no defaults at all when there aren't any reasonable ones}.
Software should also \textbf{check that configuration values are reasonable at startup}:
few things are as annoying as having a program announce after running for two hours
that it isn't going to save its results
because the requested directory doesn't exist.

@@

In order to be useful, this test script must be easy to find and run. 
Many build systems will also run unit tests if provided them at compile time.
If the build system is not amenable to testing, a 
single file in the project's root directory named \texttt{runtests.sh}
or something equally obvious is a much better solution than documenting
test cases and requiring people to copy and paste them into the shell.

Equally, the test script's output must be easy to interpret. Screens
full of correlation coefficients do not qualify: instead, the script's
output should be simple to understand for non-experts,
such as one line per test, with the test's name
and its pass/fail status, followed by a single summary line saying how
many tests were run and how many passed or failed. If many or all tests
fail because of missing dependencies, that fact should be displayed
once, clearly, rather than once per test, so that users have a clear
idea of what they need to fix and how much work it's likely to take.

Finally,
studies have found that the ease with which people can start making
contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
By making it simpler for outsiders to contribute,
a test suite of any kind also makes it more likely that they will, and software
with collaborators stands a better chance of surviving in the busy field of
scientific software.

@@

Given a set of parameters and a dataset, a particular version of a program
should produce the same results every time it is run
to aid testing, debugging, and reproducibility.
Even minor changes to code can cause minor changes in output because of floating-point issues,
which means that getting exactly the same output for the same input and parameters
probably won't work during development,
but it should still be a goal for people who have deployed a specific version.

@@

Given the size of biological data, it is unreasonable to suggest that
random algorithms be removed. However, most programs use a pseudo-random
number generator, which uses a starting seed and an equation to
approximate random numbers. Setting the seed to a consistent value
can remove randomness between runs. Allow the user to optionally provide
the seed as an input parameter, thus rendering the program deterministic
for those cases where it matters. If the seed is set internally (e.g.,
using clock time), echo it to the output for re-use later. 
If setting the seed is not possible, make sure the acceptable tolerance is
known and detailed in documentation and in the tests.

@@

\item[\textbf{Documentation}] We think it is very important for
  developers to document their work, but our experience is that people
  are unlikely do it during normal development, and we don't want to
  recommend practices that people won't actually adopt. However, it is worth
  noting that software that is widely used and contributed to has and enforces
  the need for good documentation~\cite{gentleman2004}.

\item[\textbf{Build Systems}] 
\end{description}

>>>>

@@

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. The lack of
robustness also leads to many duplicated efforts,
slowing the pace of research~\cite{prabhu2011,lawlor2015}.
Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
hundreds of tools to perform the same task. In the Bioinformatics
Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.
Some of these tools are legitimate efforts to improve the state-of-the-art, but
often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
after publication~\cite{nekrutenko2012}.

@@

\textbf{List required dependencies:} Often, software depends on
specific versions of libraries, modules, or operating systems. 
This is
entirely reasonable as long as it is properly documented. Include the name and
version number for each dependency.  Several methods exist for acquiring
required packages: either provide the commands necessary to download the
dependencies; link to the software homepage; or (our preference) use a
dependency manager. We address dependencies in more detail in rule 5.

\textbf{Provide compile/installation instructions:} If the software needs to be compiled
or installed, list those instructions in the README. Also
mention if you recommend they use a pre-compiled binary instead
through a system such as pip, yum or apt.

\textbf{List input and output files}: All possible input and output files
should be listed in this section. Do the files conform to a particular
standard, an extension of an existing format, or is it your own
format? If using a standard format, link to the specification and
version. If you extend the standard or have your own format, define it
explicitly, listing all the required fields and acceptable values.
(You get bonus points if you include a script to convert between
a standard format and your file format). If there is no rigorous format
(such as with log files), show an example file, or the first few lines,
and explain what the sections mean.

Input files and their formats are included in most documentation, but
intermediate, auxiliary and log files are often missing.
\emph{All} files should be listed in the README, even those considered self-explanatory. Log
files are often full of valuable information that can be
mined for the user's specific purpose. If your users might need to know,
``Does this program report the percentage of reads trimmed to remove
adapter sequences?'' they should be able to check the README and confidently
say, ``Yes, it is in the log file''.

\textbf{State attributions and licensing:} Attributions are how you credit
your main contributors; licenses are how you want others to use and
credit your software. READMEs are usually available \emph{before}
the software is installed (or even downloaded).
Leave no
question in anyone's mind about whether your software can be used
commercially, how much modification is permitted, and how other software
needs to credit you. If your software is not open source, state that clearly.
Attributions can also contain a list of ``expert'' users
that can be contacted if new users have problems with the software.

@@

Usage should provide all of the information necessary to run the
software. It is invoked either by running the software without
any arguments; running the software with incorrect arguments; or by
explicitly choosing a help or usage option.
More standard command-line behaviours are detailed in \cite{Seemann2013}.

@@

\textbf{The version of your software should be easily available by 
supplying \texttt{-\/-version} or \texttt{-v} on the command line}. This command should print
the software name and version number, and it should
also be \textbf{included in all of the program's output}, particularly debugging
traces: if someone needs help, it's important that they be able to tell
whoever's helping them which version of the software they're using.

While new releases may make a program better in general,
they can simultaneously create work for someone
who integrated the old version into their own workflow a year or two ago,
and won't see any benefits from upgrading.
A program's authors should therefore \textbf{ensure that every released version
continues to be available.}
A number of mechanisms exist for
controlled release that range from as simple as adding an appropriate
commit message or tag to version control, to official releases alongside
code on Bitbucket or GitHub, to depositing into a
repository like apt, yum, homebrew, CPAN, etc. Choose the method that
best suits the number and expertise of users you anticipate.

@@

Unfortunately, reusing software introduces dependencies, which can
bring their own special pain.  The interface between two software
packages can also be a source of considerable frustration: all too
often, support requests descend into debugging errors produced by the
other project. 
In addition, every package someone has to install
before being able to use yours is a
possible (some would say ``likely'') source of frustration for some
potential user. The way the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explicitly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example, GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

@@

Second, \textbf{ensure the appropriate software and version is available}.
Early in execution, check whether the dependency is executable and whether the
version is compatible.
Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using a
dependency management system (see Rule 5).

@@

\section{Use a build utility and package manager.}

In all but the simplest of scripts, \textbf{we highly recommend using a build
automation utility to compile code and/or manage your software workflow}. The benefits of
using a build utility like Make, Rake, Maven, Ant or MS Build vastly outweighs the
drawbacks associated with the learning curve. Most importantly, a build system
lets someone else build your software the same way you would. Most build
utilities can also be configured to run tests during every build and
re-generate documentation, and can be hooked into continuous integration,
cutting releases, and deploying the software to other environments.

Build utilities rely on any dependencies being present in the environment for a
successful build, and so developers should
\textbf{document \emph{all} dependencies, preferably in a machine-readable form}.
Package managers like apt and yum are available on most *nix systems, and
application package managers exist for specific languages like Python (pip),
Java (Maven/Gradle), and Ruby (RubyGems). These package managers can be used
together with the build utility to ensure that dependencies are available at
compile/run time.
For example, it is common for Python projects to include a file called
\texttt{requirements.txt} that lists the names of required libraries,
along with version ranges:

@@

This file can be read by pip package manager, which can check that the
required software is available, and install it if it is not. 
Whatever is used,
we recommend that developers should \emph{always} install dependencies
using this mechanism, especially on their personal machines, so that
they're sure it works.

Conversely, developers should
\textbf{avoid depending on scripts and tools which are not available as packages}.
In many cases, a program's author may not realize that some tool was built locally, and
doesn't exist elsewhere. At present, the only sure way to discover such
unknown dependencies is to install on a system administered by someone
else and see what breaks. As use of virtualization containers becomes more
widespread, software installation can also be tested on a virtual machine or
container system like Docker.

@@

Testing the ability to install software has traditionally been regarded as difficult,
since it necessarily alters the machine on which the test is conducted.
Lightweight virtualization containers like Docker make this much easier as well, 
or simply \textbf{ask another person to try and build your software}.

@@

Providing parameters on the
command line increases the flexibility and usability of the program. You
may have determined early in development that 0.58 is an optimal seed for your
original dataset, but that doesn't mean that is the best seed for every
case. Being able to change parameters on the fly to determine if and how
they change the results is important as your software gains more users,
facilitating exploratory analysis and parameter sweeping.
\textbf{Choose reasonable defaults where they exist},
and \textbf{set no defaults at all when there aren't any reasonable ones}.

You can set reasonable default values
as long as any command line arguments
override those values. Configuration files should be used in preference
to hard-coding the defaults directly. Configuration files can be in a standard
location, e.g. \texttt{.packagerc} in the user's home directory or
provided on the command line as an additional argument.
Only values that are unlikely to
change between runs belong in the config file, such as dependencies,
servers, version numbers, network drives, and any other defaults for
your lab or institution. 
Use a default configuration file and specify all other parameters on the
command line.

When the software starts, it should \textbf{echo all parameters and software
versions to standard out or a log file alongside the results}. This
feature supports greater reproducibility because any result can be
replicated with only the previous output files as reference.
Finally,
if the software is configurable in any way,
it should \textbf{check that all input values are in a reasonable range at startup}. 
Few things are as annoying as having a program announce after running for two hours
that it isn't going to save its results
because the requested directory doesn't exist.

@@

In order to be useful, \textbf{this test script must be easy to find and run}. 
Many build systems will also run unit tests if provided them at compile time.
If the build system is not amenable to testing, a 
single file in the project's root directory named \texttt{runtests.sh}
or something equally obvious is a much better solution than documenting
test cases and requiring people to copy and paste them into the shell.

Equally, \textbf{the test script's output must be easy to interpret}. Screens
full of correlation coefficients do not qualify: instead, the script's
output should be simple to understand for non-experts,
such as one line per test, with the test's name
and its pass/fail status, followed by a single summary line saying how
many tests were run and how many passed or failed. If many or all tests
fail because of missing dependencies, that fact should be displayed
once, clearly, rather than once per test, so that users have a clear
idea of what they need to fix and how much work it's likely to take.

Studies have found that the ease with which people can start making
contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
By making it simpler for outsiders to contribute,
a test suite of any kind also makes it more likely that they will, and software
with collaborators stands a better chance of surviving in the busy field of
scientific software.

@@

Given a set of parameters and a dataset, \textbf{a particular version of a program
should produce the same results every time it is run}
to aid testing, debugging, and reproducibility.
Even minor changes to code can cause minor changes in output because of floating-point issues,
which means that getting exactly the same output for the same input and parameters
probably won't work during development,
but it should still be a goal for people who have deployed a specific version.

@@

Given the size of biological data, it is unreasonable to suggest that
random algorithms be removed. However, most programs use a pseudo-random
number generator, which uses a starting seed and an equation to
approximate random numbers. Setting the seed to a consistent value
can remove randomness between runs. \textbf{Allow the user to optionally provide
the random seed as an input parameter}, thus rendering the program deterministic
for those cases where it matters. If the seed is set internally (e.g.,
using clock time), echo it to the output for re-use later. 
If setting the seed is not possible, \textbf{make sure the acceptable tolerance is
known and detailed in documentation and in the tests}.

@@

\item[\textbf{Documentation}] We think it is very important for
  developers to document their work, but our experience is that people
  are unlikely do it during normal development, and we don't want to
  recommend practices that people won't actually adopt. However, it is worth
  noting that software that is widely used and contributed to has and enforces
  the need for good documentation~\cite{gentleman2004}.
  Numerous resources exist for writing high quality
  documentation~\cite{karimzadeh2016} and so we do not cover it here.

\end{description}
